\documentclass[11pt]{article}
\usepackage{hyperref,amsmath,bm,apacite}

\begin{document}

{\large
\begin{center}
\noindent {\em Title:} Score-based tests of measurment invariance: Use
in practice\\ \ \\

\noindent {\em Authors:} Ting Wang, Edgar Merkle, Achim Zeileis
\end{center}}

\pagestyle{empty}

% Paragraph 1: Background
%% General statement about measurement invariance


%% Summary of Merkle + Zeileis 2013

%% The tests have had only limited application to measurement
%% invariance, and their properties as applied to measurement
%% invariance are not fully known.

As the Research Topic indicates, measurement invariance is a critical
issue in psychometric modeling.  Recently, Merkle and
Zeileis \citeyear{MerZei13} studied a novel family of
score-based tests of 
measurement invariance for factor analysis models.  This family can
be used to test for measurement invariance w.r.t.\ a continuous
auxiliary variable, and the family is advantageous because it can
identify subgroups of observations and model parameters that violate
measurement invariance.  Merkle, Fan, and Zeileis \citeyear{MerFanZei}
extended this framework to testing measurement invariance w.r.t.\ an
ordinal 
auxiliary variable, yielding test statistics that are sensitive to
violations that are monotonically related to the auxiliary variable
(and insensitive to non-monotonic violations).

% Paragraph 2: Tutorial in lavaan
The papers cited in the above paragraph have been relatively technical
and slanted towards psychometricians.  In the proposed paper for the
Research Topic, we plan to focus on the 
test statistics' use in practice.  We will first provide a 
minimally-technical overview of the tests, along with a tutorial on
carrying out the tests using the R packages lavaan (for model
estimation) and strucchange (for measurement invariance testing).
These packages currently have all the functionality necessary for
carrying out the tests on arbitrary structural equation models, so the
tutorial will be especially useful for researchers who wish to apply
the tests to their own data.
%% Ting: I think the following details are too specific for the abstract.
% To carry out these tests, a confirmatory factor analysis model is fit to the data first. Then R package lavaan could
% provide casewise derivatives and observed information matrix, and then they are used to calculate the cumulative
% score process. In the end, by using different aggregation function of cumulative
% score process, we obtain three tests' statistics and \emph{p}-values, which could indicate whether or not measurement invariance is 
% violated.

% Paragraph 3: Simulations to examine issues that are of importance in
% practice.
Following the tutorial, we plan to conduct and report on
some novel simulations that are relevant to the tests' use in practice.  In
Simulation 1, we will study the extent to which the proposed
tests attribute measurement invariance violations to the correct model
parameters.  In Simulation 2, we will study the extent to which the
tests are robust to model misspecification.  Taken together, the
tutorial and simulations will provide researchers with both the means
to carry out the tests and the knowledge of when the tests are likely to
be useful in practice. 
%% Ting: For this proposed abstract, I think we do not need to discuss
%% any results.
% The resutls indicate that the tests
% could correctly locate the violation parameter, as long as the violation parameter is included
% in the model. If not, the tests show violation in more than one parameter associated with the
% particular manifest variable which is unduely modeled.
\bibliographystyle{apacite}
\bibliography{refs}


\end{document}
